### 2023-11-27 main.tex
''
^-^

\section{Introduction} 
Boundary-driven particle systems are paradigmatic models for non-equilibrium statistical mechanics. In a boundary-driven system, the model is put in contact with reservoirs at its ``boundaries'' imposing different particle densities, and a current is generated through the system. In the long time limit, a non-equilibrium steady state sets in, with a stationary value of the current. The aim of this paper is to study the mathematical aspects of the stationary measure of boundary-driven {\em multispecies} particle systems \cite{schutz2017kardar, vanicat2017exact, crampe2016integrable, finn2018matrix, floreani2022switching, casini2022uphill, casini2023density, redig2022ergodic, redig2023equilibrium, chatterjee2023multi, zhou2021orthogonal, franceschini2022orthogonal}.

@@

The remainder of this paper is organized as follows. In Section \ref{sec1} we define the open multispecies stirring
process and provide its Lie algebraic description. In Section \ref{sectionDuality} we prove duality (Theorem \ref{thm-duality}) 
and we show that, even in the presence of multispecies, the problem of characterizing the non-equilibrium 
steady state can be reduced to the computation of the absorption probabilities. Duality is proved when the 
system has a general spatial structure described by a general graph. In Section \ref{sectionIntegrabiliy} we then obtain a closed formula 
for the stationary correlations (Theorem \ref{thm-correlations}) for the case of a one-dimensional chain with at most one 
particle per site, which is driven out-of-equilbrium by two reservoirs acting at the left and right boundaries. 
Here the computation of the absorption probabilities relies on the previously formulated Matrix Product Ansatz \cite{vanicat2017exact},
which is in fact also simplified by the use of duality. Finally in Section \ref{sec5} we discuss a few extensions 
and generalizations of the model, including instantaneous thermalization limits and reaction-diffusions.

In closing this introduction we mention that we are currently working on the boundary-driven multispecies 
model with non-compact Lie-algebras, both integrable and non-integrable, which will be presented in a 
future paper.

@@

In this paper we will often use the fact that the Markov generator of the multispecies stirring process can be described in terms a Lie algebra of rank $N$.
In this section we provide details about this.

@@

\section{Duality}\label{sectionDuality}
In this section we prove the first main result of this paper. We show that the multispecies stirring process with open boundary
can be studied by means of a dual process. After recalling the definition of duality between two
Markov processes in Section \ref{def-duality}, we formulate our duality result  in Theorem \ref{thm-duality}. A proof of the theorem is given in Section \ref{proof-th-duality}.
The duality result is important since it yields several simplifications, the main one being that the
dual process has absorbing sites. Therefore, as shown in Corollary \ref{Corolollary-ABS_Corr-abstract}, the study of correlations in the non-equilibrium steady state of the multispecies stirring process with open boundary
is reduced to the study of the absorption probabilities of dual particles.

@@

	&\mathcal{P}_{x_{1},x_{2},x_{3}}(0,1,0)=\frac{x_1 (L (-x_3)+L+x_2 (x_3-2)+1)}{L(L^2-1)}\\

@@

\begin{equation}
	p_{a}p_{b}=p_{h_{c}(a)}p_{h_{c}(b)}
\end{equation}
By the arbitrariness of $a,b,$ it is true if and only if
\begin{equation}\label{p-equal}
	p_{a}=p\qquad \forall a\in \{1,\ldots,N-1\}
\end{equation}
while $p_{N}=1-\sum_{a=1}^{N-1}p_{a}$\\
For the boundary generators $\mathcal{L}_{\text{left}}^{rd}$ and $\mathcal{L}_{\text{right}}^{rd}$ we have that, analogously to what happens for the multispecies stirring process, the reservoirs parameters must be such that $\beta_{a}=\alpha_{a}$ and, due to \eqref{p-equal}, we also need $\alpha_{a}=\alpha$ for all $a\in\{1,\ldots,N-1\}$. 
\end{comment}
\subsubsection{Duality for the reaction-diffusion process}
In this section we formulate duality for the multi-species reaction diffusion process. This answer to the second question addressed at the beginning of Section \ref{Subsection-RD}.  The dual process is defined on the enlarged state space $\widetilde{\Omega}_{L}$.  
\begin{proposition}[Duality for the reaction-diffusion process]\label{propositin-duality-RD}
The reaction-diffusion multi-species stirring process $(\bm{n}(t))_{t\geq 0}$, on the state space $\Omega_{L}$, with generator $\mathcal{L}^{rd}$ defined in \eqref{RDGenerator} is dual to the process $(\bm{\xi}(t))_{t\geq 0}$ on the state space $\widetilde{\Omega}_{L}$ with dual generator
\begin{equation}\label{DualGeneratorRD}
	\widetilde{\mathcal{L}}_{\text{left}}^{rd}=\widetilde{\mathcal{L}}^{rd}+\sum_{x=1}^{L-1}\mathcal{L}_{x,x+1}^{rd}+\widetilde{\mathcal{L}}_{\text{right}}^{rd}
\end{equation}
where 
$\mathcal{L}_{x,x+1}^{rd}$ is defined in \eqref{edgeGeneratorRD} and, $\widetilde{\mathcal{L}}_{\text{left}}^{rd}$ and $\widetilde{\mathcal{L}}_{\text{right}}^{rd}$ are the absorbing dual generators $\widetilde{\mathcal{L}}_{x}$ defined in \eqref{siteDualGenerator} and acting at sites $1$ and $L$, with the specific choice of the reservoirs parameters \eqref{boundaryParamRD}. The duality function
is the same of the stirring process, i.e.
\begin{equation}
	D(\bm{n},\bm{\xi})=\prod_{a=1}^{N-1}\left(\rho_{a}^{\text{left}}\right)^{\xi_{a}^{0}}\left(\prod_{x=1}^{L}\frac{(\nu -\sum_{a=1}^{N-1}\xi_{a}^{x})!}{\nu!}\prod_{a=1}^{N-1}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}\right)\prod_{a=1}^{N-1}\left(\rho_{a}^{\text{right}}\right)^{\xi_{a}^{L+1}}
\end{equation}
(cf. \eqref{dualityElements}).
\end{proposition} 
%\begin{equation}
%	\widetilde{\Omega}=\mathbb{N}_{0}^{N}\times  \bigtimes_{x\in V} \Omega_{x}\times \mathbb{N}_{0}^{N}
%\end{equation}
%We associate an ``extra site'', denoted by $0$ and $L+1$ attached to $1$ and $L$ respectively,
%where dual particles will accumulate in the course of time. 
%and we denote its configurations by $\bm{\xi} \in \widetilde\Omega$.
%\begin{equation}
%	\bm{\xi}=\left(\xi_{1}^{0},\ldots,\xi_{N}^{0}\right)\otimes\left(\xi_{1}^{x},\ldots,\xi_{N}^{x}\right)^{\otimes L}\otimes \left(\xi_{1}^{L+1},\ldots,\xi_{N}^{L+1}\right)
%\end{equation}
%where the component $\xi_{a}^{x}$ is interpreted as the number of dual particles of type $a\in \{1,\ldots,N\}$ at site $x$, 
%and the component $\xi_{a}^{0},\xi_{a}^{L+1}$  give the number at the extra sites. \\
%where we denote the \textit{density} of the species $a\in \{2,\ldots,N\}$ imposed by the reservoir at site $x\in V$  by 
%\begin{equation}
%	\rho_{a}^{x}:=\frac{\alpha_{a}^{x}}{|\alpha^{x}|}
%\end{equation}

@@

We multiplying it by $\left(\frac{\alpha_{N}}{|\alpha|}\right)^{-(r_{1}+\ldots+r_{N-1})+(r_{1}+\ldots+r_{N-1})}$ and we obtain

>>>>

\section{Introduction} 
Boundary-driven particle systems are paradigmatic models for non-equilibrium statistical mechanics. In a boundary-driven system, the model is put in contact with reservoirs at its ``boundaries'' imposing different particle densities, and a current is generated through the system. In the long time limit, a non-equilibrium steady state sets in, with a stationary value of the current. The aim of this paper is to study the mathematical aspects of the stationary measure of boundary-driven {\em multi-species} particle systems \cite{schutz2017kardar, vanicat2017exact, crampe2016integrable, finn2018matrix, floreani2022switching, casini2022uphill, casini2023density, redig2022ergodic, redig2023equilibrium, chatterjee2023multi, zhou2021orthogonal, franceschini2022orthogonal}.

@@

The remainder of this paper is organized as follows. In Section \ref{sec1} we define the open multi-species stirring
process and provide its Lie algebraic description. In Section \ref{sectionDuality} we prove duality (Theorem \ref{thm-duality}) 
and we show that, even in the presence of multi-species, the problem of characterizing the non-equilibrium 
steady state can be reduced to the computation of the absorption probabilities. Duality is proved when the 
system has a general spatial structure described by a general graph. In Section \ref{sectionIntegrabiliy} we then obtain a closed formula 
for the stationary correlations (Theorem \ref{thm-correlations}) for the case of a one-dimensional chain with at most one 
particle per site, which is driven out-of-equilbrium by two reservoirs acting at the left and right boundaries. 
Here the computation of the absorption probabilities relies on the previously formulated Matrix Product Ansatz \cite{vanicat2017exact},
which is in fact also simplified by the use of duality. Finally in Section \ref{sec5} we discuss a few extensions 
and generalizations of the model, including instantaneous thermalization limits and reaction-diffusions.

In closing this introduction we mention that we are currently working on the boundary-driven multi-species 
model with non-compact Lie-algebras, both integrable and non-integrable, which will be presented in a 
future paper.

@@

In this paper we will often use the fact that the Markov generator of the multi-species stirring process can be described in terms a Lie algebra of rank $N$.
In this section we provide details about this.

@@

\section{Duality}\label{sectionDuality}
In this section we prove the first main result of this paper. We show that the multi-species stirring process with open boundary
can be studied by means of a dual process. After recalling the definition of duality between two
Markov processes in Section \ref{def-duality}, we formulate our duality result  in Theorem \ref{thm-duality}. A proof of the theorem is given in Section \ref{proof-th-duality}.
The duality result is important since it yields several simplifications, the main one being that the
dual process has absorbing sites. Therefore, as shown in Corollary \ref{Corolollary-ABS_Corr-abstract}, the study of correlations in the non-equilibrium steady state of the multi-species stirring process with open boundary
is reduced to the study of the absorption probabilities of dual particles.

@@

	&\mathcal{P}_{x_{1},x_{2},x_{3}}(0,1,0)=\frac{x_1 (L(1-x_3)+x_2 (x_3-2)+1)}{L(L^2-1)}\\

@@

\begin{equation}
	p_{a}p_{b}=p_{h_{c}(a)}p_{h_{c}(b)}
\end{equation}
By the arbitrariness of $a,b,$ it is true if and only if
\begin{equation}\label{p-equal}
	p_{a}=p\qquad \forall a\in \{1,\ldots,N-1\}
\end{equation}
while $p_{N}=1-\sum_{a=1}^{N-1}p_{a}$\\
For the boundary generators $\mathcal{L}_{\text{left}}^{rd}$ and $\mathcal{L}_{\text{right}}^{rd}$ we have that, analogously to what happens for the multi-species stirring process, the reservoirs parameters must be such that $\beta_{a}=\alpha_{a}$ and, due to \eqref{p-equal}, we also need $\alpha_{a}=\alpha$ for all $a\in\{1,\ldots,N-1\}$. 
\end{comment}
\subsubsection{Duality for the reaction-diffusion process}
In this section we formulate duality for the multi-species reaction diffusion process. This answer to the second question addressed at the beginning of Section \ref{Subsection-RD}.  The dual process is defined on the enlarged state space $\widetilde{\Omega}_{L}$.  
\begin{proposition}[Duality for the reaction-diffusion process]\label{propositin-duality-RD}
The reaction-diffusion multi-species stirring process $(\bm{n}(t))_{t\geq 0}$, on the state space $\Omega_{L}$, with generator $\mathcal{L}^{rd}$ defined in \eqref{RDGenerator} is dual to the process $(\bm{\xi}(t))_{t\geq 0}$ on the state space $\widetilde{\Omega}_{L}$ with dual generator
\begin{equation}\label{DualGeneratorRD}
	\widetilde{\mathcal{L}}_{\text{left}}^{rd}=\widetilde{\mathcal{L}}^{rd}+\sum_{x=1}^{L-1}\mathcal{L}_{x,x+1}^{rd}+\widetilde{\mathcal{L}}_{\text{right}}^{rd}
\end{equation}
where 
$\mathcal{L}_{x,x+1}^{rd}$ is defined in \eqref{edgeGeneratorRD} and, $\widetilde{\mathcal{L}}_{\text{left}}^{rd}$ and $\widetilde{\mathcal{L}}_{\text{right}}^{rd}$ are the absorbing dual generators $\widetilde{\mathcal{L}}_{x}$ defined in \eqref{siteDualGenerator} and acting at sites $1$ and $L$, with the specific choice of the reservoirs parameters \eqref{boundaryParamRD}. The duality function
is the same of the stirring process, i.e.
\begin{equation}
	D(\bm{n},\bm{\xi})=\prod_{a=1}^{N-1}\left(\rho_{a}^{\text{left}}\right)^{\xi_{a}^{0}}\left(\prod_{x=1}^{L}\frac{(\nu -\sum_{a=1}^{N-1}\xi_{a}^{x})!}{\nu!}\prod_{a=1}^{N-1}\frac{n_{a}^{x}!}{(n_{a}^{x}-\xi_{a}^{x})!}\right)\prod_{a=1}^{N-1}\left(\rho_{a}^{\text{right}}\right)^{\xi_{a}^{L+1}}
\end{equation}
(cf. \eqref{dualityElements}).
\end{proposition} 
%\begin{equation}
%	\widetilde{\Omega}=\mathbb{N}_{0}^{N}\times  \bigtimes_{x\in V} \Omega_{x}\times \mathbb{N}_{0}^{N}
%\end{equation}
%We associate an ``extra site'', denoted by $0$ and $L+1$ attached to $1$ and $L$ respectively,
%where dual particles will accumulate in the course of time. 
%and we denote its configurations by $\bm{\xi} \in \widetilde\Omega$.
%\begin{equation}
%	\bm{\xi}=\left(\xi_{1}^{0},\ldots,\xi_{N}^{0}\right)\otimes\left(\xi_{1}^{x},\ldots,\xi_{N}^{x}\right)^{\otimes L}\otimes \left(\xi_{1}^{L+1},\ldots,\xi_{N}^{L+1}\right)
%\end{equation}
%where the component $\xi_{a}^{x}$ is interpreted as the number of dual particles of type $a\in \{1,\ldots,N\}$ at site $x$, 
%and the component $\xi_{a}^{0},\xi_{a}^{L+1}$  give the number at the extra sites. \\
%where we denote the \textit{density} of the species $a\in \{2,\ldots,N\}$ imposed by the reservoir at site $x\in V$  by 
%\begin{equation}
%	\rho_{a}^{x}:=\frac{\alpha_{a}^{x}}{|\alpha^{x}|}
%\end{equation}

@@

We multiplying it by $\prod_{k=1}^{N-1}\left(\frac{\alpha_{N}}{|\alpha|}\right)^{-r_{k}+r_{k}}$ and we obtain
